---
  `---
title: "KEY STONE"
author: "Tio"
format: html
editor: visual
---

# Housekeeping

```{r housekeeping}
#| message: true
#| warning: false
rm(list = ls())
```

# Packages

```{r}
library(tidyverse) # Inc dplyr, ggplot2, lubridate etc
library(readxl) # For reading in excel files
```

# Load files

# Main dataset

The catch data is split across 4 tabs in one excel file Load indivudally

```{r}
catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 1)
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)

catch_data <- bind_rows(catch1, catch2, catch3, catch4)

```

# Inspect individual tabs

```{r}
catch1
catch2
catch3
catch4
```

# Bind rows together

```{r}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
# Clean up the environment
rm(catch1, catch2, catch3, catch4) # Clean up the environment
```

# Examine data

```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

Can't tell from these outputs how many levels of site and species we have. Two options: let's check those explicitly by using distinct() or can convert to factors Let's look at the distinct values first, as we might need to clean before converting to factor vars.

```{r}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```

OK, so we have four species in there as well as four sites which look to be located at different points on the Ross River. Need to: - Fix site names and place in reasonable order. - Can decide this now or we can load the metadata and ensure consistency - Fix species names - to do this we'll need to make consistent and then use the taxonomic dictionary to translate to scientific names.

# Load metadata

```{r}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
```

# Inspect

```{r}
head(metadata)
glimpse(metadata)
```

Site names are all over the place in the metadata too - will need to standardise.

```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

# Recheck

```{r}
catch_data |> 
  str()
# catch_data |> 
#   distinct(site)
glimpse(metadata)

```

# Load Taxonomic Dictionary

```{r}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
head(species_dict)
```

# Load sensor data,

```{r}
sensor_raw <- read_csv("../data/workshop2/estuary_sonde_data.csv")
```

# Check sensor data

```{r}
head(sensor_raw)
str(sensor_raw)
glimpse(sensor_raw)
summary(sensor_raw)
```

#### 
