---
title: "Workshop_1"
author: "Agustinus Satrio Supoyo"
format: html
editor: visual
---

## Introduction

This workshop provides an introduction to data manipulation and analysis in R using the `tidyverse` suite of packages. The focus is on importing data from various file formats, inspecting dataset structure, selecting relevant variables, filtering observations, sorting data, creating new variables, summarising data by groups, and visualising patterns in ecological datasets.

The `palmerpenguins` dataset is used to demonstrate these concepts and techniques.

## Session Memory Housekeeping

### Inspect Objects in the Current R Session

The `objects()` function lists all objects currently stored in the Global Environment.

```{r inspect-session-objects}
objects()
```

### Remove All Existing Objects

The following command removes all objects from the Global Environment.

```{r clear-global-environment}
rm(list = ls())
```

### Confirm That the Global Environment Is Empty

```{r confirm-empty-environment}
objects()
```

## Troubleshooting Tip

Objects may sometimes reappear in the Environment after restarting R because a hidden `.RData` file is automatically restoring data from a previous session.

The following command permanently deletes the `.RData` file from the home directory. It is shown for reference but is not run automatically during knitting.

```{r remove-rdata-file, eval=FALSE}
unlink("~/.RData")
```

This helps maintain reproducible workflows and prevents hidden objects from influencing the current analysis.

## Importing Data from Multiple File Formats

Environmental and ecological datasets are commonly stored in different file formats depending on their source. In this exercise, data are imported from three common formats: CSV, TSV, and Excel.

The `here()` package is used to create reproducible file paths relative to the project directory. This allows the workflow to run on different computers without manually changing file locations.

### Load Required Packages

```{r load-import-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(here)
```

### Practice Import A: CSV File

CSV (Comma-Separated Values) files store data as plain text, with columns separated by commas. This is one of the most widely used formats for scientific data exchange.

```{r import-csv-file}
benthic_cover <- read_csv(
  here::here(
    "data",
    "workshop1",
    "reef_cover_log.csv"
  )
)
```

### Practice Import B: TSV File

TSV (Tab-Separated Values) files use tab characters to separate columns. They are commonly generated by automated sensors, telemetry systems, and environmental monitoring equipment.

```{r import-tsv-file}
acoustic_stream <- read_tsv(
  here::here(
    "data",
    "workshop1",
    "acoustic_telemetry_stream.txt"
  )
)
```

### Practice Import C: Excel Spreadsheet

Excel files can contain multiple worksheets, formatting information, and metadata. The `read_excel()` function allows a specific worksheet to be selected during import.

```{r import-excel-file}
fisheries_annual <- read_excel(
  here::here(
    "data",
    "workshop1",
    "fish_catch_data.xlsx"
  ),
  sheet = "Commercial_2026"
)
```

## Handling Files with Non-Standard Headers

Some datasets contain notes or comments before the actual column headers. These extra rows can cause R to incorrectly identify variable names when importing data.

### Initial Import

```{r import-raw-mangrove-data}
mangrove_data <- read_csv(
  file = here::here(
    "data",
    "workshop1",
    "mangrove_survey_raw.csv"
  )
)
```

### Skip Metadata Rows and Define Missing Values

The `skip` argument removes the first five lines of field notes. The `na` argument converts known missing-value codes into proper `NA` values.

```{r import-clean-mangrove-data}
mangrove_data <- read_csv(
  file = here::here(
    "data",
    "workshop1",
    "mangrove_survey_raw.csv"
  ),
  skip = 5,
  na = c(
    ".",
    "NA",
    "9999",
    "ND",
    "blank"
  )
)
```

This ensures that the dataset is imported with the correct variable names and that missing values are properly recognised. This supports more accurate data cleaning and analysis in later steps.

## Tibbles Versus Data Frames

Datasets imported using tidyverse functions are usually stored as tibbles, which are a modern version of traditional R data frames.

Tibbles provide cleaner console output, display variable types automatically, and reduce accidental partial column matching.

### Check and Convert the Data Structure

```{r compare-data-classes}
class(benthic_cover)

benthic_cover_df <- as.data.frame(
  benthic_cover
)

class(benthic_cover_df)
```

### Compare Printed Output

```{r compare-tibble-data-frame-output}
# Print the traditional data frame
print(benthic_cover_df)

# Print the tibble
print(benthic_cover)
```

## Dataset Inspection and Structure Assessment

Before conducting any analysis, it is important to inspect the structure of a newly imported dataset. This step helps identify the available variables, their data types, dataset dimensions, and any potential issues that may require cleaning before analysis.

### Exploring the Palmer Penguins Dataset

The `palmerpenguins` dataset is used as an ecological example for practising data wrangling and exploratory analysis.

Before performing any analysis, the structure of the dataset should be inspected to identify the available variables, data types, and overall dataset dimensions.

### Install the Package

Run this command only once in the Console if the package is not already installed.

```{r install-palmerpenguins, eval=FALSE}
install.packages("palmerpenguins")
```

### Load the Package and Dataset

```{r load-palmerpenguins, message=FALSE, warning=FALSE}
library(palmerpenguins)

data("penguins")
```

### Examine the Dataset Structure

The `glimpse()` function is a tidyverse function that provides a concise overview of the dataset, including variable names, data types, and a preview of the observations.

The `str()` function is a base R function that provides a more detailed description of the dataset structure, including the number of observations, variables, and the data type of each variable.

Both functions are useful when first examining a new dataset.

```{r inspect-penguins-structure}
glimpse(penguins) # Tidyverse version

str(penguins) # Base R version
```

### Generate an Exploratory Summary

The `summary()` function provides descriptive summaries for each variable in the dataset.

```{r summarise-penguins}
summary(penguins)
```

## Selecting Relevant Variables

The `select()` function is used to retain variables that are relevant to a specific ecological question.

This simplifies downstream analysis by removing unnecessary columns and keeping only the required information.

### Select Morphometric Variables by Name

```{r select-morphometric-variables}
morphology_metrics <- select(
  penguins,
  species,
  bill_length_mm,
  bill_depth_mm,
  body_mass_g
)

glimpse(morphology_metrics)
```

### Select a Continuous Block of Variables

The colon operator `:` can be used inside `select()` to retain a continuous range of variables.

```{r}
spatial_block <- select(penguins, species:island)
```

### Remove Unnecessary Variables

The minus sign `-` removes selected variables while retaining all remaining columns.


```{r}
clean_scientific_fields <- select(penguins, -year)
```

## Filtering Observations

The `filter()` function is used to subset a dataset according to specific conditions. This allows the analysis to focus on relevant observations, such as a particular species or individuals that meet a numerical threshold.

### Filter a Single Species

```{r}
adelie_cohort <- filter(penguins, species == "Adelie")
```

### Filter Using a Numerical Threshold

The following code retains only penguins with a body mass greater than 4,500 grams.

```{r filter-heavy-penguins}
heavy_penguins <- filter(penguins, body_mass_g > 4500)
```

### Filter Using Multiple Conditions

The `&` operator combines multiple conditions. The following code retains Gentoo penguins sampled on Biscoe Island.


```{r}
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")
```

### Filter Using a Set of Values

The `%in%` operator retains observations that match any value in a specified set.

```{r}
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))
```

## Sorting and Ordering Data

The `arrange()` function was used to sort the dataset based on one or more variables. This allows for easy identification of patterns, trends, or outliers by ordering the data in a meaningful way.

### Sort penguins by ascending body mass (Default setting: Smallest mass first)

```{r}
lightest_first <- arrange(penguins, body_mass_g)
print(lightest_first)
```

### Sort penguins in descending sequence using the desc() layout wrapper

```{r}
heaviest_first <- arrange(penguins, desc(body_mass_g))
print (heaviest_first)
```

### Execute nested sorting criteria: Group by species, then sort by descending bill length

```{r}
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
print(stratified_morphology)
```

## Chaining Operations with the Pipe

The pipe operator (`%>%`) was used to chain multiple data manipulation steps together in a clear and readable format. This allows for a more intuitive workflow by connecting operations sequentially without the need for intermediate variables.

```{r}
penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
print(penguins_final)
```

## Creating New Variables with `mutate()`

The `mutate()` function was used to create new variables based on existing ones. This allows for the derivation of additional insights or metrics that may be relevant for analysis.

### Calculate a new morphological ratio in our environment

```{r}
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )
print(penguin_ratios)
```

### View your newly engineered variables appended to the far-right columns

```{r}
glimpse(penguin_ratios)
```

## Data Aggregation and Ecological Summarisation

The `group_by()` and `summarise()` functions were used to aggregate data and calculate summary statistics for specific groups within the dataset. This allows for the analysis of patterns and trends across different categories or levels of a variable.

### Grouping our active memory penguins by species

```{r}
grouped_penguins <- group_by(penguins, species)
```

### Notice that the table looks identical, but metadata notes 'Groups: species \[3\]'

```{r}
print(grouped_penguins)
```

### Calculate Mean Body Mass by Species

```{r}
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)
```

## Handling Missing Values During Aggregation

When calculating summary statistics, missing values (NA) can affect the results. The `na.rm = TRUE` argument was used within summary functions to exclude NA values from calculations, ensuring that the summary statistics are accurate and not skewed by missing data.

###Overcoming the missing value trap using na.rm = TRUE
```{r}

biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal)

```

## Wrangling and Plotting in Parallel

This section summarises and visualises penguin body mass across species and islands. \# Summary table: mean body mass by species and island

```{r}
body_mass_summary <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_body_mass_g = mean(body_mass_g, na.rm = TRUE),
    sample_size = n(),
    .groups = "drop"
  )

body_mass_summary

```

### Boxplot of Body Mass by Species and Island

```{r}
penguins |>
  ggplot(
    aes(
      x = species,
      y = body_mass_g,
      fill = island
    )
  ) +
  geom_boxplot() +
  labs(
    title = "Penguin Body Mass by Species and Island",
    x = "Species",
    y = "Body mass (g)",
    fill = "Island"
  ) +
  theme_minimal()
```

or

### Faceted Boxplot by Island

```{r}
penguins |>
  ggplot(
    aes(
      x = species,
      y = body_mass_g
    )
  ) +
  geom_boxplot() +
  facet_wrap(~ island) +
  labs(
    title = "Penguin Body Mass by Species Across Islands",
    x = "Species",
    y = "Body mass (g)"
  ) +
  theme_minimal()
```

## Piping Directly to Visualisation

This section pipes the summarised penguin body mass data directly into `ggplot2`. This workflow is useful for quickly visualising ecological patterns without creating unnecessary intermediate objects.

```{r}
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot

```

### Interpretation

The graph shows that Gentoo penguins have the highest average body mass, whereas Adelie and Chinstrap penguins have lower and relatively similar body masses. A statistical test would likely indicate significant differences in body mass among species, with species identity having a stronger influence than island location.

##challenge

###Change standard deviation with standard error of the mean (SEM) in the error bars. Standar Deviation (SD)

```{r}
penguin_summary <- penguins %>%
  group_by(species) %>%
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n()
  )
print(penguin_summary)
```

###Standard Error of the Mean (SEM)

```{r}
penguin_summary <- penguins %>%
  group_by(species) %>%
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    se_mass = sd_mass / sqrt(n)
  )
print(penguin_summary)
```

### Plot Mean Body Mass with Standard Error
```{r}
ggplot(
  penguin_summary,
  aes(x = species,
      y = mean_mass)
) +
  geom_col() +
  geom_errorbar(
    aes(
      ymin = mean_mass - se_mass,
      ymax = mean_mass + se_mass
    ),
    width = 0.2
  )
```

### Interpretation

Gentoo penguins exhibited the highest mean body mass, whereas Adelie penguins were the lightest species. Replacing standard deviation with standard error resulted in noticeably smaller error bars, indicating greater precision around the estimated species means. This suggests that although individual body masses vary within species, the mean body mass estimates are relatively reliable. The observed differences likely reflect species-specific ecological adaptations and resource use strategies.

## Creating Output Folders

```{r create-output-folders}
if (!dir.exists("outputs/figures")) {
  dir.create(
    "outputs/figures",
    recursive = TRUE
  )
}

if (!dir.exists("outputs/tables")) {
  dir.create(
    "outputs/tables",
    recursive = TRUE
  )
}

if (!dir.exists("outputs/rdata")) {
  dir.create(
    "outputs/rdata",
    recursive = TRUE
  )
}
```

## Exporting Summary Tables

The summarised biological data are exported as a CSV file.


```{r}
write_csv(
  biological_signal,
  here::here(
    "outputs",
    "penguin_species_mass_summary.csv"
  )
)
```


## Saving a Cleaned Dataset as an RDS File

A cleaned subset of the Palmer Penguins dataset is created by retaining key morphological variables and removing incomplete observations.

An RDS file preserves R-specific data types and can be loaded efficiently in future analyses.
 
### Create a cleaned penguin dataset
    ```{r}
   
    clean_cohort <- penguins %>%
      select(
        species,
        bill_length_mm,
        bill_depth_mm,
        flipper_length_mm,
        body_mass_g,
        sex
      ) %>%
      drop_na()


    ```

### Saving our cleaned penguin morphology cohort as a native R binary file
```{r}

saveRDS(
  clean_cohort,
  here::here(
    "outputs",
    "clean_penguin_morphology_cohort.rds"
  )
)
```

## Exporting Figures

The final visualisation was exported as a high-resolution PNG image using \`ggsave()\`. Saving figures ensures that outputs can be reused in reports, presentations, and future analyses without recreating the plot.

```{r}
ggsave(
  filename = here::here(
    "outputs",
    "mass_compare_plot.png"
  ),
  plot = mass_compare_plot,
     
  width = 120,
  height = 120,
  units = "mm",
  dpi = 300
)
```
