---
title: "Workshop_1"
author: "Agustinus Satrio Supoyo"
format: html
editor: visual
---

##Introduction

This workshop provides an introduction to data manipulation and analysis in R using the tidyverse suite of packages. The focus is on importing data from various file formats, inspecting dataset structure, selecting relevant variables, filtering observations, sorting data, creating new variables, summarising data by groups, and visualising patterns in ecological datasets. The Palmer Penguins dataset is used as an example to demonstrate these concepts and techniques.

## Session Memory Housekeeping

## Inventory every active object currently residing in session RAM

```{r}
objects()
```

## Remove All Existing Objects

```{r}
rm(list = ls())
```

## Confirm that global session memory now completely vacant

```{r}
objects()
```

##Troubleshooting Tip: ##If objects appear in the Environment despite starting a new R session, a hidden `.RData` file may be automatically restoring data from previous sessions. To remove it and ensure a clean working environment, run:

```{r}
unlink("~/.RData")
```

##This helps maintain reproducible workflows and prevents hidden objects from influencing current analyses .

## Importing Data from Multiple File Formats

Environmental and ecological datasets are commonly stored in different file formats depending on their source. In this exercise, data were imported from three common formats: CSV, TSV, and Excel spreadsheets. The `here()` package was used to create reproducible file paths relative to the project directory, ensuring that the workflow can be executed on any computer without modifying file locations.

### Load Required Packages

```{r}
library(tidyverse)
library(readxl)
```

### Practice Import A: CSV File

CSV (Comma-Separated Values) files store data as plain text with columns separated by commas. This is one of the most widely used formats for scientific data exchange.

```{r}
benthic_cover <- read_csv(
  here::here("data", "workshop1", "reef_cover_log.csv")
)
```

### Practice Import B: TSV File

TSV (Tab-Separated Values) files use tab characters to separate columns and are commonly generated by automated sensors, telemetry systems, and environmental monitoring equipment.

```{r}
acoustic_stream <-read_tsv(
  here::here("data", "workshop1", "acoustic_telemetry_stream.txt")
)
```

### Practice Import C: Excel Spreadsheet

Excel files can contain multiple worksheets, formatting information, and metadata. The `read_excel()` function allows a specific worksheet to be targeted during import.

```{r}
fisheries_annual <- read_excel(
  here::here("data", "workshop1", "fish_catch_data.xlsx"),
  sheet = "Commercial_2026"
)
```

## Handling Files with Non-Standard Headers

Some datasets contain notes or comments before the actual column headers. These extra rows can cause R to incorrectly identify variable names when importing data.

```{r}
mangrove_data <- read_csv(file = here::here("data", "workshop1","mangrove_survey_raw.csv"))
```

# Use args within read_csv to skip headers and declare missing flags

```{r}
mangrove_data <- read_csv(
  here::here("data","workshop1", "mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA
```

## This ensures that the dataset is imported with correct variable names and that missing values are properly recognized, facilitating accurate data analysis and cleaning in subsequent steps.

## Tibbles versus Data Frames

Tibbles Datasets imported using tidyverse functions are stored as tibbles, a modern version of the traditional R data frame. Tibbles provide cleaner console output, display variable types automatically, and prevent accidental partial column matching, making them safer and easier to work with during data analysis.

```{r}
class(benthic_cover)

benthic_cover_df <- as.data.frame(benthic_cover)

class(benthic_cover_df)
```

VS Data Frames

```{r}
# Print the old-style dataframe structure to view
print(benthic_cover_df)
# And compare with tibble alternative
print(benthic_cover)

```

## Dataset Inspection and Structure Assessment

Before conducting any analysis, it is important to inspect the structure of a newly imported dataset. This step helps identify the available variables, their data types, dataset dimensions, and any potential issues that may require cleaning prior to analysis. \# Exploring the Palmer Penguins Dataset The Palmer Penguins dataset was loaded as an example ecological dataset for practicing data wrangling and exploratory analysis. Before performing any analysis, the structure of the dataset was inspected to identify the available variables, data types, and overall dataset dimensions. \# Install the data package (execute this command once in your console pane and then delete!)

```{r}
#install.packages("palmerpenguins")
```

# Load the package data into active memory

```{r}
library(palmerpenguins)
data("penguins")
```

Examine the structure of the dataset - always do this when loading a new dataset!

glimpse() is a tidyverse function that provides a concise overview of the dataset, showing variable names, data types, and a preview of the data.

str() is a base R function that gives a more detailed structure of the dataset, including the number of observations and variables, as well as the data types of each variable. Both functions are useful for understanding the dataset before performing any analysis or cleaning steps.

```{r}
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version
```

# Generate an exploratory summary matrix

```{r}
summary(penguins)
```

## Selecting Relevant Variables

The `select()` function was used to isolate variables relevant to specific ecological questions. This allows unnecessary columns to be removed and simplifies downstream analysis by retaining only the required information. \# Vertically slice specific morphometric variables by explicit name

```{r}
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)

```

# Retain a continuous block of attributes using the colon operator

```{r}
spatial_block <- select(penguins, species:island)
```

# Discard logistics tracking attributes while preserving everything else using the minus sign

```{r}
clean_scientific_fields <- select(penguins, -year)
```

## Filtering Observations

The `filter()` function was used to subset the dataset based on specific criteria, such as selecting only certain species or observations that meet particular conditions. This allows for focused analysis on relevant subsets of the data. \# Isolate observations belonging to a single categorical target group

```{r}
adelie_cohort <- filter(penguins, species == "Adelie")
```

# Sift out individuals using continuous numerical boundary thresholds

# Preserves only large penguins whose mass exceeds 4500 grams

```{r}
heavy_penguins <- filter(penguins, body_mass_g > 4500)
```

# Combine multiple conditional parameters across separate attributes

# Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island

```{r}
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")
```

# Sift records matching multiple targeting flags within an explicit set

```{r}
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))
```

## Sorting and Ordering Data

The `arrange()` function was used to sort the dataset based on one or more variables. This allows for easy identification of patterns, trends, or outliers by ordering the data in a meaningful way.

# Sort penguins by ascending body mass (Default setting: Smallest mass first)

```{r}
lightest_first <- arrange(penguins, body_mass_g)
print(lightest_first)
```

# Sort penguins in descending sequence using the desc() layout wrapper

```{r}
heaviest_first <- arrange(penguins, desc(body_mass_g))
print (heaviest_first)
```

# Execute nested sorting criteria: Group by species, then sort by descending bill length

```{r}
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
print(stratified_morphology)
```

## Chaining Operations with the Pipe

The pipe operator (`%>%`) was used to chain multiple data manipulation steps together in a clear and readable format. This allows for a more intuitive workflow by connecting operations sequentially without the need for intermediate variables.

```{r}
penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
print(penguins_final)
```

## Creating New Variables with `mutate()`

The `mutate()` function was used to create new variables based on existing ones. This allows for the derivation of additional insights or metrics that may be relevant for analysis.

# Calculate a new morphological ratio in our environment

```{r}
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )
print(penguin_ratios)
```

# View your newly engineered variables appended to the far-right columns

```{r}
glimpse(penguin_ratios)
```

## Data Aggregation and Ecological Summarisation

The `group_by()` and `summarise()` functions were used to aggregate data and calculate summary statistics for specific groups within the dataset. This allows for the analysis of patterns and trends across different categories or levels of a variable.

# Grouping our active memory penguins by species

```{r}
grouped_penguins <- group_by(penguins, species)
```

# Notice that the table looks identical, but metadata notes 'Groups: species \[3\]'

```{r}
print(grouped_penguins)
```

# Collapsing the buckets into explicit summary metrics

```{r}
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)
```

## Handling Missing Values During Aggregation

When calculating summary statistics, missing values (NA) can affect the results. The `na.rm = TRUE` argument was used within summary functions to exclude NA values from calculations, ensuring that the summary statistics are accurate and not skewed by missing data.

```{r}
# Overcoming the missing value trap using na.rm = TRUE
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal)

```

## Wrangling and Plotting in Parallel

This section summarises and visualises penguin body mass across species and islands. \# Summary table: mean body mass by species and island

```{r}
body_mass_summary <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_body_mass_g = mean(body_mass_g, na.rm = TRUE),
    sample_size = n(),
    .groups = "drop"
  )

body_mass_summary

```

Visualisation: Boxplot of body mass by species and island

```{r}
penguins |>
  ggplot(
    aes(
      x = species,
      y = body_mass_g,
      fill = island
    )
  ) +
  geom_boxplot() +
  labs(
    title = "Penguin Body Mass by Species and Island",
    x = "Species",
    y = "Body mass (g)",
    fill = "Island"
  ) +
  theme_minimal()
```

Or using facet_wrap for a different layout

```{r}
penguins |>
  ggplot(
    aes(
      x = species,
      y = body_mass_g
    )
  ) +
  geom_boxplot() +
  facet_wrap(~ island) +
  labs(
    title = "Penguin Body Mass by Species Across Islands",
    x = "Species",
    y = "Body mass (g)"
  ) +
  theme_minimal()
```

## Piping Directly to Visualisation

This section pipes the summarised penguin body mass data directly into `ggplot2`. This workflow is useful for quickly visualising ecological patterns without creating unnecessary intermediate objects.

```{r}
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot

```

#Interpretation:

The graph shows that Gentoo penguins have the highest average body mass, whereas Adelie and Chinstrap penguins have lower and relatively similar body masses. A statistical test would likely indicate significant differences in body mass among species, with species identity having a stronger influence than island location.

#challenge

#Change standard deviation with standard error of the mean (SEM) in the error bars. Standar Deviation (SD)

```{r}
penguin_summary <- penguins %>%
  group_by(species) %>%
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n()
  )
print(penguin_summary)
```

Standard Error of the Mean (SEM)

```{r}
penguin_summary <- penguins %>%
  group_by(species) %>%
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    se_mass = sd_mass / sqrt(n)
  )
print(penguin_summary)
```

```{r}
ggplot(
  penguin_summary,
  aes(x = species,
      y = mean_mass)
) +
  geom_col() +
  geom_errorbar(
    aes(
      ymin = mean_mass - se_mass,
      ymax = mean_mass + se_mass
    ),
    width = 0.2
  )
```

#Interpretation

Gentoo penguins exhibited the highest mean body mass, whereas Adelie penguins were the lightest species. Replacing standard deviation with standard error resulted in noticeably smaller error bars, indicating greater precision around the estimated species means. This suggests that although individual body masses vary within species, the mean body mass estimates are relatively reliable. The observed differences likely reflect species-specific ecological adaptations and resource use strategies.

#creates output folder

```{r}
if (!dir.exists("outputs/figures"))
  dir.create("outputs/figures")

if (!dir.exists("outputs/tables"))
  dir.create("outputs/tables")

if (!dir.exists("Rdata"))
  dir.create("Rdata")
```

#Save the table as a CSV file 1. Exporting our collapsed summary table as a universal flat text file

```{r}
write_csv(
  biological_signal,
  here::here(
    "outputs",
    "penguin_species_mass_summary.csv"
  )
)
```

2.  Saving our cleaned morphological cohort table as a native R binary file

    ##Saving a Cleaned Dataset as an RDS File\
    \
    A cleaned subset of the Palmer Penguins dataset was created by retaining key morphological variables and removing incomplete observations. The resulting dataset was saved as an RDS file to preserve data types and allow efficient reuse in future analyses.

    ```{r}
    # Create a cleaned penguin dataset
    clean_cohort <- penguins %>%
      select(
        species,
        bill_length_mm,
        bill_depth_mm,
        flipper_length_mm,
        body_mass_g,
        sex
      ) %>%
      drop_na()


    ```

```{r}
# Saving our cleaned penguin morphology cohort as a native R binary file
saveRDS(
  clean_cohort,
  here::here(
    "outputs",
    "clean_penguin_morphology_cohort.rds"
  )
)
```

\## Exporting Figures\
\
The final visualisation was exported as a high-resolution PNG image using \`ggsave()\`. Saving figures ensures that outputs can be reused in reports, presentations, and future analyses without recreating the plot.

```{r}
ggsave(
  filename = here::here(
    "outputs",
    "mass_compare_plot.png"
  ),
  plot = mass_compare_plot,+
     
  width = 120,
  height = 120,
  units = "mm",
  dpi = 300
)
```
