---
title: "R for Marine Science Workshop 2"
author: "Ben Cresswell"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

# Title

## Subtitle??

### Section heading

#### What's this

### Introduction

This qmd file covers all the material for the second R for Marine Science workshop. Here we are going to cover: - Tidying data - Pivoting wider and longer - Spearating and uniting variables - Dealing with strings and dates - Joining separate dataframes/tibbles - Handling missing data values

```{r setup, include=FALSE}
# Set options
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

### Load packages and housekeeping

```{r housekeeping}
rm(list = ls()) # Clear out environment
library(tidyverse)
```

### Tidy data using tidyverse

#### Look at inbuilt datasets

##### Table 1 - tidy

```{r}
table1 <- table1
```

##### Table 2 - cases and population combined into type

```{r}
table2 <- table2
```

##### Table 3

```{r}
table3 <- table3
```

##### Inspect together

```{r}
table1
table2
table3
```

# Things we can do with tidy data (table1)

## Compute rates

```{r}
table1 %>% 
  mutate(rate = cases / population * 10000)
```

## Compute cases per year

```{r}
table1 %>% 
  count(year, wt = cases)
```

# Visualise changes over time

```{r fig.width=150/25.4, fig.height=100/25.4}
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country), size = 5)
```

# Pivoting

Sometimes we need to reshape data to make it tidy. This is often the case when we have multiple measurements of the same variable, but they are stored in separate columns. We can use the `pivot_longer()` function from the `tidyr` package to reshape data from wide to long format, and the `pivot_wider()` function to reshape data from long to wide format.

```{r}
billboard
```

This is a wide dataset - multiple measurements of the same variable (rank) are stored in separate columns (wk1, wk2, wk3, etc.). We can use `pivot_longer()` to reshape this data into a tidy format.

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank")
```

```{r}
getwd
library(here)
reef_cover <- read_csv(here::here("data/workshop1/reef_cover_log.csv"))
reef_cover <- read_csv("../data/workshop1/reef_cover_log.csv")

head(reef_cover)
print(reef_cover)

print("Ben")
head("Ben")
```

```{r}
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "General Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)



```

```{r}
# Joining metadata to our observations
joined_data <- observations |>
  left_join(site_metadata, by = join_by(site_code))


joined_data <- observations |>
  right_join(site_metadata, by = join_by(site_code))


print(joined_data)


```




```{r}
# Make sure packages are loaded
library(tidyverse)
library(palmerpenguins)

# The messy metadata provided by your colleague
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

print(island_metadata)


```

# Need to fix island_name so that anything with lower-case at start becomes upper case
```{r}
#island_metadata <- 
  island_metadata |>
  mutate(island_name = str_to_title(str_trim(island_name)))
```



```{r}
# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
#clean_sites <- 
  messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Replace any spaces with underscores
    
    #site_clean = str_trim(site_clean),
    
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_")
    # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
   
  )


```









